---
output:
  word_document: default
  html_document: default
---
## Divvy Bikeshare Full Year Analysis
The purpose of this script is to consolidate downloaded Divvy data into a single dataframe and then conduct simple analysis to help answer the key question: “In what ways do members and guest riders use Divvy bikes differently?”

#### Loading libraries
```{r}
library(tidyverse)  
library(lubridate)  
library(ggplot2)

```

#### Data Collection
reading .csv files in to dataframes
```{r}
tripdata_2020_09 <- read_csv("202009-divvy-tripdata.csv")
tripdata_2020_10 <- read_csv("202010-divvy-tripdata.csv")
tripdata_2020_11 <- read_csv("202011-divvy-tripdata.csv")
tripdata_2020_12 <- read_csv("202012-divvy-tripdata.csv")
tripdata_2021_01 <- read_csv("202101-divvy-tripdata.csv")
tripdata_2021_02 <- read_csv("202102-divvy-tripdata.csv")
tripdata_2021_03 <- read_csv("202103-divvy-tripdata.csv")
tripdata_2021_04 <- read_csv("202104-divvy-tripdata.csv")
tripdata_2021_05 <- read_csv("202105-divvy-tripdata.csv")
tripdata_2021_06 <- read_csv("202106-divvy-tripdata.csv")
tripdata_2021_07 <- read_csv("202107-divvy-tripdata.csv")
tripdata_2021_08 <- read_csv("202108-divvy-tripdata.csv")
```
#### Data consolidation
First a look at the column names of each dataframe
```{r}
colnames(tripdata_2020_09)
colnames(tripdata_2020_10)
colnames(tripdata_2020_11)
colnames(tripdata_2020_12)
colnames(tripdata_2021_01)
colnames(tripdata_2021_02)
colnames(tripdata_2021_03)
colnames(tripdata_2021_04)
colnames(tripdata_2021_05)
colnames(tripdata_2021_06)
colnames(tripdata_2021_07)
colnames(tripdata_2021_08)
```

Now to check dataframe structures for incongruencies
```{r}
str(tripdata_2020_09)
str(tripdata_2020_10)
str(tripdata_2020_11)
str(tripdata_2020_12)
str(tripdata_2021_01)
str(tripdata_2021_02)
str(tripdata_2021_03)
str(tripdata_2021_04)
str(tripdata_2021_05)
str(tripdata_2021_06)
str(tripdata_2021_07)
str(tripdata_2021_08)
```
Converting column data types to match for consolidation
```{r}
tripdata_2020_09 <- mutate(tripdata_2020_09, start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
tripdata_2020_10 <- mutate(tripdata_2020_10, start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
tripdata_2020_11 <- mutate(tripdata_2020_11, start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
```

Combine dataframes into one

```{r}
all_data <- bind_rows(tripdata_2020_09, tripdata_2020_10, tripdata_2020_11, tripdata_2020_12, tripdata_2021_01, tripdata_2021_02, tripdata_2021_03, tripdata_2021_04, tripdata_2021_05, tripdata_2021_06, tripdata_2021_07, tripdata_2021_08)
```

Removing columns unnecessary for analysis
```{r}
all_data <- all_data %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng))

```
 
#### Data cleaning and prep

first a brief overall look at the data and it's properties
```{r}
colnames(all_data)  #List of column names
nrow(all_data)  #How many rows are in data frame?
dim(all_data)  #Dimensions of the data frame?
head(all_data)  #See the first 6 rows of data frame.  Also tail(all_trips)
str(all_data)  #See list of columns and data types (numeric, character, etc)
summary(all_data)  #Statistical summary of data. Mainly for numerics

```

There are a few issues that needs to be fixed:

(1) In the "member_casual" column, there are two names for members ("member" and "Subscriber") and two names for casual riders ("Customer" and "casual"). These will be consolidated from four to two labels

(2) The data can only be aggregated at the ride-level, which is too granular. Additional columns of data -- such as day, month, year -- will be added to provide additional opportunities to aggregate the data.

(3)  A calculation will be needed to determine the length of each trip, this will add another column which will be called "ride_length" to the data frame. 


Reassigning values to the correct labels: 
```{r}
all_data <-  all_data %>% 
  mutate(member_casual = recode(member_casual
                           ,"Subscriber" = "member"
                           ,"Customer" = "casual"))

```

Adding additional columns for detailed analysis
```{r}
all_data$date <- as.Date(all_data$started_at) #The default format is yyyy-mm-dd
all_data$month <- format(as.Date(all_data$date), "%m")
all_data$day <- format(as.Date(all_data$date), "%d")
all_data$year <- format(as.Date(all_data$date), "%Y")
all_data$day_of_week <- format(as.Date(all_data$date), "%A")
```

Adding a function to calculate the ride length
```{r}
all_data$ride_length <- difftime(all_data$ended_at,all_data$started_at)

```

inspecting the structure of the dataframe
```{r}
str(all_data)
```

Converting ride_length to numeric data type to perform calculations
```{r}
is.factor(all_data$ride_length)
all_data$ride_length <- as.numeric(as.character(all_data$ride_length))
is.numeric(all_data$ride_length)

```

Remove bad data: A few hundred times bikes where checked out for service and repairs by "HQ QR" while some rides show a negative ride time.
A new database called "all_data_v2" will be created by filtering out such entries
```{r}
all_data_v2 <- all_data[!(all_data$start_station_name == "HQ QR" | all_data$ride_length<0),]

```

#### Descriptive Analysis

Descriptive analysis on ride_length (all figures in seconds)
```{r}
aggregate(all_data_v2$ride_length ~ all_data_v2$member_casual, FUN = mean)
aggregate(all_data_v2$ride_length ~ all_data_v2$member_casual, FUN = median)
aggregate(all_data_v2$ride_length ~ all_data_v2$member_casual, FUN = max)
aggregate(all_data_v2$ride_length ~ all_data_v2$member_casual, FUN = min)

```

Now a look at the average ride time by day for members vs casual users
```{r}
aggregate(all_data_v2$ride_length ~ all_data_v2$member_casual + all_data_v2$day_of_week, FUN = mean)

```

Fixing order of days in dataframe
```{r}
all_data_v2$day_of_week <- ordered(all_data_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

```

Now recalculating average ride time by day for member vs casual users
```{r}
aggregate(all_data_v2$ride_length ~ all_data_v2$member_casual + all_data_v2$day_of_week, FUN = mean)

```

Analyze rider data by user type and week day
```{r}
all_data_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  
  group_by(member_casual, weekday) %>%  		
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 		
  arrange(member_casual, weekday)			

```
#### Visualizations

Visualization of number rides by user type 
```{r}
all_data_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")+
  scale_y_continuous(labels = function(x) paste0(x/1e3, " Thousand"))+
  labs(title = "Daily rides by usertype", x = "Day of the week", y = "Number of rides", fill = "user type")

```

Visualization of average ride duration(seconds)
```{r}
all_data_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(),average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Average daily ride duration by user type", x ="Day of the week", y = "Average ride duration", fill = "User type")

```

